---
title: "Airline Delay Prediction Proposal"
author: "Logistics Transport Aviation"
subtitle: "1423543"
date: "30/08/2021"
output:
  html_document:
  number_sections: yes
fig_caption: yes
toc: yes
toc_depth: '2' 
---
  
```{r setup, include=FALSE}
## knitr::opts_chunk$set(echo = TRUE)
## install.packages("kableExtra")
library(tidyverse)
library(kableExtra)
```

## Airline Delay Prediction

Aircraft delays have serious economic impacts that represent a logistical headache for airlines, airports, airline crew, and passengers alike. In a 2010 study by the National Center of Excellence For Aviation Operations Research, an estimated 31.2 billion US dollars was lost due to direct and indirect costs of airline delays in the United States for 2007. These expenses included maintenance costs, extra paid hours for flight crew, the extra use of fuel, and airport fees. As an example, airlines have dedicated time slots at airport gates and unexpected delays exceeding the expected time slot can cost thousands of dollars by the minute.

Many factors can contribute to aircraft delays. These may include unavoidable reasons such as inclement weather, unforeseen aircraft mechanical breakdowns, or a multitude of passenger-related incidents. Airline delays then create complicated scenarios for airline schedulers as a delay for one aircraft may impact the flight schedule for all consecutive flights using that aircraft. Passengers with connecting flights may need to be ticketed on other flights, reserve flight crew may need to be called if the current flight crew are going to exceed their flight hours, and delays may continue to have a snowball effect on later flights causing them to also be delayed (Ball, 2010).

Thus, understanding the impact of these variables on aircraft on-time performance can help elucidate ways and methods to mitigate these airline delays. By doing so, steps can be made to reduce airline costs associated with these delays. We identify the primary stakeholders to be investors and shareholders for the financial performance of the airline. In addition, airline passengers are important stakeholders as delays negatively influence repurchase intentions and sentiment across word-of-mouth (Kim, 2016).

Herein, we propose an exploratory analysis and prediction model on airline delays utilizing data from the United States in 2019.

## Stakeholders

Throughout the project, we'll need to manage our external and internal stakeholders in our data science project.

```{r stakeholder_analysis, echo=FALSE}
text_tbl <- data.frame(
  Stakeholders = c("Airline company executives", "Airline company employees", "Shareholders and investors of Airline companies", "Consumers of airline companies", "Suppliers for Airline companies", "Government"),
  Description = c(
    "The main participants in the aviation industry. 
This project can help them improve their operational planning and execution and any related products and services. ",
    "Pilots, engineers, and flight attendants ensure that passengers' needs are appropriately met and satisfied.  ", 
    "Their primary role is to provide capital in the form of share capital. Airlines performance may affect their investing behaviours. ", "Airline company/industry’s primary revenue source. Customers are the most important factor in this industry; without customers, airlines would not exist. ", "Form in which how much the company orders supplies in impacts inventory, e.g Aircraft manufacters, fuel and food/beverage resources. ", "Government and policy makers can impact how the industry will be regulated, such as climate change, tourism, and infrastructure policy.  "
  )
)

kbl(text_tbl) %>%
  kable_paper(full_width = F) %>%
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2, width = "30em")
```


## Datasets

Main dataset is from The United States Department of Transportation, it is a comprehensive dataset with delay time, reasons for delays, airline code, origin and destination airport code. To improve the readibility of the dataset, we included a lookup table that has information on airline codes and airline names. Furthermore, the airport name corresponding to airport codes are retrieved through API, then it is merged to the main dataset. More datasets, such as Skytrax Airline review which records sentiment will be considered and might be used for merging in later stages. 

## Regression modelling techniques

One regression model we are considering using in this report is a multivariate linear regression (MLR) model.
As flight delays could have a multitude of causes, including multiple reasons per flight, a MLR would be most suitable for this model. Some causes could be weather, operating issues, air traffic, and seasonal variabilities such as holidays. We will intend to examine how much influence each variable has on the dependent variable.



## Impact analysis

There are few potential issues that may arise in this project.

- The data could corrupt while doing the project
- Inaccurate statistical analysis and cannot improve the analysis
- Potential to choose the wrong visualizations to display the results
- Conveying the results to stakeholders and transferring wrong information 

# Conclusion/Summary



# References

Ball, M. (2010, November). Total Delay Impact Study A Comprehensive Assessment of the Costs and Impacts of Flight Delay in the United States. https://isr.umd.edu/NEXTOR/pubs/TDI_Report_Final_11_03_10.pdf


# Appendex

```{r add_packages, message=FALSE, warning=FALSE}
  library(psych)
  library(devtools)

  library(tidyverse)
  library(dplyr)
  library(ggplot2)
  library(Amelia)

  library(knitr)
  library(kfigr)
```
=======

API call 
```{r}
library(httr)
library(jsonlite)
library(dplyr)

icao <- GET(
  url = 'https://applications.icao.int/dataservices/api/safety-characteristics-list?api_key=8d00ef90-0982-11ec-9d72-8160549d64ab&airports=&states=USA'
)
response <- content(icao, 'parsed')
API_data <- fromJSON(response)

#Only pick those columns that are useful later on 
API_data <- API_data %>% select(airportCode, airportName)
```
On time data 
```{r}
ontime_raw <- read.csv("https://raw.githubusercontent.com/charlottetse33/36103_Logistics-Transport-Aviation/main/test%20on%20time%20report.csv")
```
Airline DOT_ID correspond to airline name data
```{r}
#Seperate loading data to prevent R from crashing
id_airline <- read.csv("https://raw.githubusercontent.com/charlottetse33/36103_Logistics-Transport-Aviation/main/dot_id_airline.csv")

```
Add actual airline names to on time dataset 
```{r}
colnames(id_airline) <- c("DOT_ID_Reporting_Airline", "Airline")
  
ontime_1 <- left_join(ontime_raw, id_airline, by = "DOT_ID_Reporting_Airline")

```
Add API Airport name 
```{r}
ontime_2 <- ontime_1 %>% mutate(Origin_airportCode = paste("K",Origin, sep = ""), Dest_airportCode = paste("K",Dest, sep = ""))

Originjoint <- left_join(ontime_2, API_data, by = c("Origin_airportCode" = "airportCode"))
names(Originjoint)[names(Originjoint) == "airportName"] <- "Origin_airportName"

ontime_final <- left_join(Originjoint, API_data, by = c("Dest_airportCode" = "airportCode"))
names(ontime_final)[names(ontime_final) == "airportName"] <- "Dest_airportName"


#New column names from merging data: 
#Airline, Origin_airportName, Dest_airportName

```

