---
title: "Airline Delay Prediction Proposal"
author: "Logistics Transport Aviation"
subtitle: "1423543"
date: "30/08/2021"
output:
  html_document:
  number_sections: yes
fig_caption: yes
toc: yes
toc_depth: '2' 
---
  
```{r setup, include=FALSE}
## knitr::opts_chunk$set(echo = TRUE)
## install.packages("kableExtra")
library(tidyverse)
library(kableExtra)
```

## Airline Delay Prediction

Aircraft delays have serious economic impacts that represent a logistical headache for airlines, airports, airline crew, and passengers alike. In a 2010 study by the National Center of Excellence For Aviation Operations Research, an estimated 31.2 billion US dollars was lost due to direct and indirect costs of airline delays in the United States for 2007. These expenses included maintenance costs, extra paid hours for flight crew, the extra use of fuel, and airport fees. As an example, airlines have dedicated time slots at airport gates and unexpected delays exceeding the expected time slot can cost thousands of dollars by the minute.

Many factors can contribute to aircraft delays. These may include unavoidable reasons such as inclement weather, unforeseen aircraft mechanical breakdowns, or a multitude of passenger-related incidents. Airline delays then create complicated scenarios for airline schedulers as a delay for one aircraft may impact the flight schedule for all consecutive flights using that aircraft. Passengers with connecting flights may need to be ticketed on other flights, reserve flight crew may need to be called if the current flight crew are going to exceed their flight hours, and delays may continue to have a snowball effect on later flights causing them to also be delayed (Ball, 2010).

Thus, understanding the impact of these variables on aircraft on-time performance can help elucidate ways and methods to mitigate these airline delays. By doing so, steps can be made to reduce airline costs associated with these delays, benefiting the airlines' stakeholders. We identify the primary stakeholders to be investors and shareholders for the financial performance of the airline. In addition, airline passengers are important stakeholders as delays negatively influence repurchase intentions and sentiment across word-of-mouth (Kim, 2016).

Herein, we propose an exploratory analysis and prediction model on airline delays utilizing data from the United States in 2019.

## Stakeholders

```{r stakeholder_analysis, echo=FALSE}
text_tbl <- data.frame(
  Stakeholders = c("Airline company executives", "Airline company employees", "Shareholders and investors of Airline companies", "Consumers of airline companies", "Suppliers for Airline companies", "Government"),
  Description = c(
    "The main participants in the aviation industry. 
This project can help them improve their operational planning and execution and any related products and services. ",
    "Pilots, engineers, and flight attendants ensure that passengers' needs are appropriately met and satisfied.  ", 
    "Their primary role is to provide capital in the form of share capital. Airlines performance may affect their investing behaviours. ", "Airline company/industry’s primary revenue source. Customers are the most important factor in this industry; without customers, airlines would not exist. ", "Form in which how much the company orders supplies in impacts inventory, e.g Aircraft manufacters, fuel and food/beverage resources. ", "Government and policy makers can impact how the industry will be regulated, such as climate change, tourism, and infrastructure policy.  "
  )
)

kbl(text_tbl) %>%
  kable_paper(full_width = F) %>%
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2, width = "30em")
```

## Research Questions

As the airline network contains complex interactions between regulatory, passenger, and airline-specific needs, predicting delays and understanding the primary causes is relatably complicated. The main question to answer revolves around which factors are most correlated to, and influence greatly, on-time aircraft performance. From here, smaller questions can be asked such as whether delays are impacted by weather, date, time of day, route, aircraft type, or economic category.

## Datasets

The main dataset is acquired through The United States Department of Transportation. It is a comprehensive dataset with delay times, airline codes, origin and destination airport codes, dates, flight time, etc. To improve the readability of the dataset, we included a lookup table that has information on airline codes and their airline names. Furthermore, the airport name corresponding to airport codes are retrieved through an API, then merged with the main dataset. More datasets, such as Skytrax Airline review which record passenger sentiment will be considered and might be used for merging in later stages. Other examined datasets included US weather events, airplane crash and fatality records, aircraft employment data, and records for airlines and their aircraft types.

## Regression modelling techniques

The regression model we are considering using in this report is a multivariate linear regression (MLR) model. As flight delays could have a multitude of causes, including multiple reasons per flight, a MLR would be most suitable for this model. Some causes could be weather, operating issues, air traffic, and seasonal variabilities such as holidays. We will intend to examine how much influence each variable has on the dependent variable.

For this model, we assume that there is a linear relationship between airline delay and the various potential causes. We also assume that the residuals are normally distributed, and that the independent variables do not correlate with each other. We also assume that the variance of error terms across all delays are equally distributed.



## Impact analysis

There are few potential issues that may arise in this project. Although the use of a web API carries the most up-to-date data, a loss of connection with the API server could make compiling data difficult. Secondly, the data itself contains millions of observations from 2019 through 2021. Computing power and memory storage are potential hurdles for analysis. In fact, one dataset was over 80GB and was unable to be loaded into the analysis program. Other potential issues may arise with merging data where a joining dataset may be missing values. Lastly, the statistical analysis relies on several assumptions specific to multivariate linear regression. Should these assumptions be broken, it may lead to difficulty in model accuracy.


# Conclusion/Summary



# References

Ball, M. (2010, November). Total Delay Impact Study A Comprehensive Assessment of the Costs and Impacts of Flight Delay in the United States. https://isr.umd.edu/NEXTOR/pubs/TDI_Report_Final_11_03_10.pdf

Kim, N. Y., & Park, J. W. (2016). A study on the impact of airline service delays on emotional reactions and customer behavior. Journal of Air Transport Management, 57, 19–25. https://doi.org/10.1016/j.jairtraman.2016.07.005

# Appendex

```{r add_packages, message=FALSE, warning=FALSE}
  library(psych)
  library(devtools)

  library(tidyverse)
  library(dplyr)
  library(ggplot2)
  library(Amelia)

  library(knitr)
  library(kfigr)
```
=======

Call web API for Airport Names 
```{r, message=FALSE, warning=FALSE}
library(httr)
library(jsonlite)
library(dplyr)

icao <- GET(
  url = 'https://applications.icao.int/dataservices/api/safety-characteristics-list?api_key=8d00ef90-0982-11ec-9d72-8160549d64ab&airports=&states=USA'
)
response <- content(icao, 'parsed')
API_data <- fromJSON(response)

#Only pick those columns that are useful later on 
API_data <- API_data %>% select(airportCode, airportName)
```
Airline On-time data 
```{r, message=FALSE, warning=FALSE}
ontime_raw <- read.csv("https://raw.githubusercontent.com/charlottetse33/36103_Logistics-Transport-Aviation/main/test%20on%20time%20report.csv")
```
Airline DOT_ID corresponding to Airline Names
```{r, message=FALSE, warning=FALSE}
#Seperate loading data to prevent R from crashing
id_airline <- read.csv("https://raw.githubusercontent.com/charlottetse33/36103_Logistics-Transport-Aviation/main/dot_id_airline.csv")

```
Merge Airline Names with On-time Dataset 
```{r, message=FALSE, warning=FALSE}
colnames(id_airline) <- c("DOT_ID_Reporting_Airline", "Airline")
  
ontime_1 <- left_join(ontime_raw, id_airline, by = "DOT_ID_Reporting_Airline")

```
Merge API Airport names with On-time Dataset 
```{r, message=FALSE, warning=FALSE}
ontime_2 <- ontime_1 %>% mutate(Origin_airportCode = paste("K",Origin, sep = ""), Dest_airportCode = paste("K",Dest, sep = ""))

Originjoint <- left_join(ontime_2, API_data, by = c("Origin_airportCode" = "airportCode"))
names(Originjoint)[names(Originjoint) == "airportName"] <- "Origin_airportName"

ontime_final <- left_join(Originjoint, API_data, by = c("Dest_airportCode" = "airportCode"))
names(ontime_final)[names(ontime_final) == "airportName"] <- "Dest_airportName"

#New column names from merging data: 
#Airline, Origin_airportName, Dest_airportName

```

