---
title: "Presentation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)
library(httr)
library(jsonlite)
library(dplyr)
library(psych)
library(devtools)
library(tidyverse)
library(ggplot2)
library(Amelia)
library(knitr)
library(googledrive)
library(purrr)
```


```{r, message=FALSE, warning=FALSE}
icao <- GET(
  url = 'https://applications.icao.int/dataservices/api/safety-characteristics-list?api_key=8d00ef90-0982-11ec-9d72-8160549d64ab&airports=&states=USA'
)
response <- content(icao, 'parsed')
API_data <- fromJSON(response)

#Only pick those columns that are useful later on 
API_data <- API_data %>% select(airportCode, airportName)
```

On time data (might change later)

Google drive
Load the below code and it will connect to your googgle drive
(Still error)
```{r}
#url <- "https://drive.google.com/drive/u/3/folders/1UO8Yn8cZqf-FAQ9Vlgsu7wN5YXY9_3uU"

drive_find(n_max = 30)
#id <- "1UO8Yn8cZqf-FAQ9Vlgsu7wN5YXY9_3uU"
#read.csv(sprintf("https://docs.google.com/uc?id=%s&export=download", id))

folder <- drive_get(id = "1UO8Yn8cZqf-FAQ9Vlgsu7wN5YXY9_3uU")
csv_files <- drive_ls(folder, type = "csv")
csv_files$id
#download <- sapply(csv_files$id, function(x) sprintf("https://docs.google.com/uc?id=%s&export=download", x))

example <- drive_get(id = csv_files$id[1])

#trial --> del []
#csv_files$id[1]

#This code does not work might be because file too large, so can't read
read_csv(paste("https://drive.google.com/uc?id=", "1pY0KMo7iM3ugcckl-Hpj7GA4JG73KyUq", sep = ""))

#sapply(download[c(1,2)], function(x) read.csv(x))
```

Load from own device 

```{r}
user_specific <- "C:/Users/tsetc/OneDrive/Desktop/dataset for logistics/20"

data <- mapply(function(x,y) read.csv(paste(user_specific,x,"_",y,".csv", sep = "")), 19:21, c('01','02','03','04'))

map_df(data, ~as.data.frame(t(.)))
```


```{r}
id_airline <- read.csv("https://raw.githubusercontent.com/timothywallaby/36103_Logistics-Transport-Aviation/main/L_UNIQUE_CARRIERS%20(1).csv")
colnames(id_airline) <- c("OP_UNIQUE_CARRIER", "Airline")
  
ontime_1 <- left_join(ontime_raw, id_airline, by = "OP_UNIQUE_CARRIER")

ontime_2 <- ontime_1 %>% mutate(Origin_airportCode = paste("K",ORIGIN, sep = ""), Dest_airportCode = paste("K",DEST, sep = ""))

Originjoint <- left_join(ontime_2, API_data, by = c("Origin_airportCode" = "airportCode"))
names(Originjoint)[names(Originjoint) == "airportName"] <- "Origin_AIRPORTNAME"

ontime_3 <- left_join(Originjoint, API_data, by = c("Dest_airportCode" = "airportCode"))
names(ontime_3)[names(ontime_3) == "airportName"] <- "Dest_AIRPORTNAME"
```


```{r}
#New column names from merging data: 
#Airline, Origin_AIRPORTNAME, Dest_AIRPORTNAME
```

Clean data 
```{r}
#If there is CANCELLATION_CODE, remove row 
# Remove those without ARR_DELAY
ontime_4 <- ontime_3[ontime_3$CANCELLED == 0,]
ontime_5 <- ontime_4[!is.na(ontime_4$ARR_DELAY),]
```

Selecting columns 
```{r}
ontime_6 <- ontime_5 %>% select(YEAR, QUARTER, MONTH, DAY_OF_MONTH, DAY_OF_WEEK,AIR_TIME,ACTUAL_ELAPSED_TIME,CARRIER_DELAY,WEATHER_DELAY
,NAS_DELAY,SECURITY_DELAY,LATE_AIRCRAFT_DELAY,FIRST_DEP_TIME,TOTAL_ADD_GTIME,LONGEST_ADD_GTIME, Airline,Origin_AIRPORTNAME, Dest_AIRPORTNAME, DEP_DELAY, ARR_DELAY)
```

Check 
```{r}
summary(ontime_6)
```


EDA
```{r}
ontime_final %>% group_by(MONTH, DAY_OF_MONTH) %>% summarise (arr_delay = mean(ARR_DELAY, na.rm = TRUE), dep_delay = mean(DEP_DELAY, na.rm = TRUE)) %>% ggplot() + geom_point(aes(x = DAY_OF_MONTH, y = arr_delay, colour = dep_delay))
#There is relationship bettween arr_delay and dep_delay
```

########################################################################################################
AFTER Presentation 

Regression 

Need to label encode categorical variables first (or we can consider one-hot encoding)
```{r}
#Airline, Origin_AIRPORTNAME, DEST_AIRPORTNAME
sapply(ontime_6, class)
```

```{r}
#change those "character" into factor then into numeric 
ontime_6$Airline <- as.numeric(as.factor(ontime_6$Airline))
ontime_6$Origin_AIRPORTNAME <- as.numeric(as.factor(ontime_6$Airline))
ontime_6$Dest_AIRPORTNAME <- as.numeric(as.factor(ontime_6$Airline))
```

```{r}
sapply(ontime_6, class)
```

Regression Model 
```{r}
glm = lm(ARR_DELAY ~ QUARTER + MONTH + DAY_OF_MONTH + DAY_OF_WEEK +AIR_TIME + ACTUAL_ELAPSED_TIME + CARRIER_DELAY + WEATHER_DELAY
+ NAS_DELAY,SECURITY_DELAY + LATE_AIRCRAFT_DELAY + FIRST_DEP_TIME + TOTAL_ADD_GTIME + LONGEST_ADD_GTIME +  Airline,Origin_AIRPORTNAME + Dest_AIRPORTNAME + DEP_DELAY, data = ontime_final)
```

VIF (to test multicollinearity)
```{r}
vif(glm)
```

Multicollinearity 
VIF, pearson's correlation test - numerical
chi-square test - categorical 
point biserial correlation test - numerical & categorical(need dummy variable) 
