---
title: "Airline Delay Prediction Report for American Airlines"
author: "Logistics Transport Aviation"
date: "03/10/2021"
output:
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    fig_caption: yes
subtitle: STDS - AT2
---
  
```{r setup, include=FALSE}
## knitr::opts_chunk$set(echo = TRUE)
## install.packages("kableExtra")
library(tidyverse)
library(kableExtra)
```

# Introduction

Aircraft delays have serious economic impacts that represent a logistical headache for airlines, airports, airline crew, and passengers alike. In a 2010 study by the National Center of Excellence For Aviation Operations Research, an estimated 31.2 billion US dollars was lost due to direct and indirect costs of airline delays in the United States for 2007. These expenses included maintenance costs, extra paid hours for flight crew, the extra use of fuel, and airport fees. As an example, airlines have dedicated time slots at airport gates and unexpected delays exceeding the expected time slot can cost thousands of dollars by the minute.

Similar to other big US airlines such as Delta and United, American Airlines is presenting loses in the amount of billions of dollars per year since COVID impacted the world. 

(Include financial figures)

Many factors can contribute to aircraft delays. These may include unavoidable reasons such as inclement weather, unforeseen aircraft mechanical breakdowns, or a multitude of passenger-related incidents. Airline delays then create complicated scenarios for airline schedulers as a delay for one aircraft may impact the flight schedule for all consecutive flights using that aircraft. Passengers with connecting flights may need to be ticketed on other flights, reserve flight crew may need to be called if the current flight crew are going to exceed their flight hours, and delays may continue to have a snowball effect on later flights causing them to also be delayed (Ball, 2010).

Thus, understanding the impact of these variables on aircraft on-time performance can help elucidate ways and methods to mitigate these airline delays. By doing so, steps can be made to reduce airline costs associated with these delays, benefiting the airlines' stakeholders. We identify the primary stakeholders to be investors and shareholders for the financial performance of the airline. In addition, airline passengers are important stakeholders as delays negatively influence repurchase intentions and sentiment across word-of-mouth (Kim, 2016).

Herein, we propose an exploratory analysis and prediction model on airline delays utilizing data from the United States in 2019. For our analysis we are going to compare American Airlines to Delta and United Airlines, based on the fact that these 3 US airlines are market leaders in terms of Brand value, Market value and Domestic Market Share. We also took into consideration that these 3 airlines cater to the same segment of customers, that is the reason why big low-cost airlines such as Southwest were not considered for this analysis. 

# Research Question

## What are the main causes for Airline delays?

When looking at the main causes of delays for American Airlines flights between 2019 and 2021 we can see that the main reason corresponds to a Late arriving aircraft, meaning that a previous flight with same aircraft arrived late, causing the present flight to depart late.  This is followed closely by the second reason, an Air carrier delay, in this case a delay or cancellation can be attributed to circumstances within the airline's control such as maintenance or crew problems, aircraft cleaning, baggage loading, fueling, among others. National Aviation system delays, Weather and security issues have a lower impact on the total amount of delays. If we go back to the first reason, which can be explained by the following four, we can assume that most of those delays were caused due to circumstance within the airline’s control as well. Meaning most of the delays can be attributed to the airlines operations and therefore we have a great opportunity to reduce costs and improve our results. 

(Include graph reasons/delays in minutes)

## How is American Airlines on time performance compare to our two main competitors?

We found that all three airlines have improved their on-time performance considerably from 2019 to 2021. It is particularly interesting to us to see how United Airlines has improved their average delay time by almost 23% in just 2 years and how Delta has kept an steady leadership on Airtime performance. This once again shows us the great opportunity we have in our hands to not only improve our On-time performance, reduce costs and improve passengers’ satisfaction, but also to keep up with our main competitors and possibly beat them by learning from their good practices and most importantly taking a proactive approach to improve our current situation. 

(Include boxplot competitor performance)

# Problem Statement

During the last couple of years around 30% of the domestic flights in the United States have presented some sort of delay, this shows us the big opportunity we have in front of us. Specially if we take into consideration that every minute of delay costs airlines around $74 dollars, and that just during 2020 American Airlines domestic flights reported over 4 million minutes in delays just in domestic flights, meaning an approximate $300 million dollars were lost for this reason. 

(Include stack bar Domestic on time performance 3 airlines)

# Exploratory Data Analysis (EDA)

Since the problem deviates based on reports, we have to look at certain metrics based on delays as the end objective.

## Simple Exploratory Data Analysis 

Thoughout the EDA, we found few items just based on the frequency 

### Metrics on data types

### Analysis on categorical variables

### Analysis on numerical variables

## Advanced EDA

### Sumarry Statistics

### Summarry Stat plots

(Inclusion of distribution of variables)

### Correlation Matrix

### Nominal, ordinal and biserial tests

#### Nominal

#### Chi-square independence test 
explain chi-square independence test 

The Chi-square independence test is a statistical hypothesis test used to compare the size of discrepancies between the observed results and expected results (University of Southampton, n.d.).
The purpose of it is to determine if there is a significant relationship between two categorical (especially nominal in nature) variables (Statisticssolutions, n.d.). Or the difference between observed results and expected results is just due to chance.
A contingency table is commonly used to examine this relationship (Statisticssolutions, n.d.). 
(do we need to add an example of contingency table to explain this test? Or use our data to explain would be more appropriate )
The formula for Chi-Square is 

(insert a pic/ I don’t know how to insert a formula into Rmarkdown)
Where: c= Degrees of freedom
O = Observed values, the actual collected data.
E = Expected values.
Sigma = added up a number of values in Chi-squared test.
The null hypothesis: there is no significant difference between the observed and expected frequencies.

https://www.statisticssolutions.com/free-resources/directory-of-statistical-analyses/chi-square/

Conducted individual chi-square test, such as 
- origin airport code and origin state abr, p-value < 2.2e-16, meaning ... 
(look up how to interpret it)
- manufacturer and airline , p-value < 2.2e-16, meaning ... 

change 

After doing a correlation plot on numerical variables of the first regression, we found some NA values among the values of origin airport, destination airport, model, and manufacturer in the result. Firstly, we thought that is due to collinearity; therefore, a Chi-square independence test was conducted as a justification to remove some categorical variables. 
The followings are the steps the Chi-square test conducted and results:
First, test the dependency between categorical variables origin_airport_code and origin_state_abr. 
Null hypothesis: origin airport and origin state are independent of each other.
Alternative hypothesis: origin airport and origin state are not independent of each other.

> chisq.test(data_chi)
          Pearson’s chi-squared test
Data: data_chi
x-squared = 27000, df = 1269, p-value < 2.2e-16

The p-value < 2.2e-16, which is close to 0. Therefore, it’s statistically significant. Reject the null hypothesis: they are independent of each other, which means there is a correlation between them. These two variables are not independent, which means there is a strong association, which will weaken the statistical power of our regression model (Jim Frost, n.d.).Therefore we removed one of them, the variable origin_state. 

Next, we examined predictor variables manufacturer and airline
The test result: 
> chisq.test(data_chi)
          Pearson’s chi-squared test
Data: data_chi
x-squared = 1675.7, df = 6, p-value < 2.2e-16

p-value is less than 2.2e-16, again like the previous result, it’s statistically significant. 
This time the variable manufacturer is removed. 


#### Point Biseral
Point biserial correlation test is for when one variable is numerical, another is dichotomous. One of the assumptions for this test is the normality of the numerical variable. In our case, none of our numerical variables are normal. We also considered doing ANOVA test to get a better sense of how categorical and numerical variables are associated but it also has the assumotion 

ANOVA assumes that the data is normally distributed. The ANOVA also assumes homogeneity of variance, which means that the variance among the groups should be approximately equal. ANOVA also assumes that the observations are independent of each other.
(https://www.statisticssolutions.com/free-resources/directory-of-statistical-analyses/anova/#:~:text=ANOVA%20assumes%20that%20the%20data,are%20independent%20of%20each%20other.)

### Further cleaning

#### Outliers 
tried isolation Forrest to detect outliers and eliminate them but the file is too large and the computer couldn’t run. (Limitation) 

also tried to remove outliers based on quatile

End up doing data trimming on both ends of the distributions of numerical variables


# Modeling

## Min-max normalization
Explain why we need min max normalization on data --> easier to interpret the result 

## Dummy variables
we use dummy variables on categorical variables instead of label encoding it. 
(Explain why one-hot encoding is better than label encoding - not ordinal variable)

Thinking about n vs n-1 cateogories 

For one hot encoding, make (n-1) new columns for n factors in categorical values
Pal session → one hot encode all columns 
https://towardsdatascience.com/one-hot-encoding-multicollinearity-and-the-dummy-variable-trap-b5840be3c41a

(Use of reference) 


## Mutivariate Linear Model

With Justification, & assumptions 

Removal of collinearity (add interaction variable)  

We attempted the first regression model by fitting in all the data, but there were approximately 300 variables which makes it difficult for us to interpret anything and most intercepts and coefficients aren't statistically significant. The R squared is very low, close to 0, even though the p-value of R-squared is smaller than 0.05. (explain R-squared)

## MV ln. model summary 

Describe coefficients & p-values 

X squared 

## MV VIF and Boruta algorithm 

Explanation of variance 

Which variables to choose  

## Modified mlm 

Mlm summary 

Variance plots (4) 

Variance descriptions 

## Results

## Analysis


# Limitations

## Dataset limitations

## Model limitations
In our analysis, we rely more on correlation test and EDA to verify the insights we get from linear regression as our confidence in the linear regression model is low. 

Our data is of time series nature, if a flight delays at 8:30 am for example, it will affect other flights delay departure time as well. There is autocorrelation between different observations, which results in the linear regression residuals not being independent. 
 -> insert residual plot
Furthermore, the response variable is skewed which violates the normality assumption. Even if we sqrt transform it, it still has heavy tails.
-> insert qq residual plot

To model time series data, linear regression is not appropriate and should not be considered if we want to do predictive analysis. Time series analysis with models such as ARIMA would be advised. However, linear regression is not completly useless, it still gives a rough idea of the relationship between variables although the result of it should be taken with caution. 

Even in time series analysis which can be conducted in later stages, when modeling intervention models, a regression model is needed. For example, the time series before the interventions was modeled by the seasonal ARIMA (1,0,0) x (0,1,0)12, a regression model is needed to describe the relationship before and after the intervention. 

## Conclusion

As the airline network contains complex interactions between regulatory, passenger, and airline-specific needs, predicting delays and understanding the primary causes is relatably complicated. 

Our main question to answer is: **Which factors are most correlated to, and influence greatly, on-time aircraft performance?**. 

From here, smaller questions can be asked such as: Are delays impacted by weather, date, time of day, route, aircraft type, or economic category?

The conclusion should reprise the questions and conclusions of the introduction, perhaps augmented by some additional observations or details gleaned from the analysis
section. New questions, future work, etc., can also be raised here.


## References

Ball, M. (2010, November). Total Delay Impact Study A Comprehensive Assessment of the Costs and Impacts of Flight Delay in the United States. https://isr.umd.edu/NEXTOR/pubs/TDI_Report_Final_11_03_10.pdf

Jim Frost,(n.d.). Multicollinearity in Regression Analysis: Problems, Detection,and Solutions. Statisctics By Jim. https://statisticsbyjim.com/regression/multicollinearity-in-regression-analysis/

Kim, N. Y., & Park, J. W. (2016). A study on the impact of airline service delays on emotional reactions and customer behavior. Journal of Air Transport Management, 57, 19–25. https://doi.org/10.1016/j.jairtraman.2016.07.005

Staticssolutions.(n.d.). Chi-Square Test of Independence. https://www.statisticssolutions.com/free-resources/directory-of-statistical-analyses/chi-square/?__cf_chl_jschl_tk__=pmd_JY1I6Untqs....RRSWJRpFhDJLifp7Aj06oAjnKFFfg-1633007460-0-gqNtZGzNAiWjcnBszQil

University of Southampton. (n.d.). Chi Square. Is there a statistically significant relationship between a student’s Year 11 truancy and his or her enrolment in full time education after secondary school? https://www.southampton.ac.uk/passs/full_time_education/bivariate_analysis/chi_square.page

## Appendix

```{r add_packages, message=FALSE, warning=FALSE}
  library(psych)
  library(devtools)

  library(tidyverse)
  library(dplyr)
  library(ggplot2)
  library(Amelia)

  library(knitr)
  
```


To get the latest content, we are using a web API call for Airport Names. The `httr::get` will assist us in fetching the content in the JSON format. API key is being used for authenticating our call with the API provider. Since JSON cannot be directly consumed by regression packages, we deploy `jsonlite::fromJSON` to convert to an R dataframe.

```{r, message=FALSE, warning=FALSE}
library(httr)
library(jsonlite)
library(dplyr)

icao <- GET(
  url = 'https://applications.icao.int/dataservices/api/safety-characteristics-list?api_key=8d00ef90-0982-11ec-9d72-8160549d64ab&airports=&states=USA'
)
response <- content(icao, 'parsed')
API_data <- fromJSON(response)

#Only pick those columns that are useful later on 
API_data <- API_data %>% select(airportCode, airportName)
```

Scrapping AIrfleet data from web 
```{r, eval = FALSE}
# Explicit URL
url <- 'https://www.planemapper.com/airlines/AAL'

AA <- url %>% 
  read_html() %>% 
  html_nodes(xpath='//*[@id="aircrafts"]') %>% 
  html_table()

AA_fleet <- AA[[1]]

View(AA_fleet)

# Write AA_fleet to feather
write_feather(AA_fleet, here('AA_fleet.feather'))

# Explicit URL
swaurl <- 'https://www.planemapper.com/airlines/SWA'

SWA <- swaurl %>% 
  read_html() %>% 
  html_nodes(xpath='//*[@id="aircrafts"]') %>% 
  html_table()

SWA_fleet <- SWA[[1]]

View(SWA_fleet)

# Write AA_fleet to feather
write_feather(SWA_fleet, here('SWA_fleet.feather'))

# Explicit URL
jbuurl <- 'https://www.planemapper.com/airlines/JBU'

JBU <- jbuurl %>% 
  read_html() %>% 
  html_nodes(xpath='//*[@id="aircrafts"]') %>% 
  html_table()

JBU_fleet <- JBU[[1]]

# Write AA_fleet to feather
write_feather(JBU_fleet, here('JBU_fleet.feather'))

# Explicit URL
dalurl <- 'https://www.planemapper.com/airlines/DAL'

DAL <- dalurl %>% 
  read_html() %>% 
  html_nodes(xpath='//*[@id="aircrafts"]') %>% 
  html_table()

DAL_fleet <- DAL[[1]]

# Write AA_fleet to feather
write_feather(DAL_fleet, here('DAL_fleet.feather'))

ualurl <- 'https://www.planemapper.com/airlines/UAL'

UAL <- ualurl %>% 
  read_html() %>% 
  html_nodes(xpath='//*[@id="aircrafts"]') %>% 
  html_table()

UAL_fleet <- UAL[[1]]

# Write AA_fleet to feather
write_feather(UAL_fleet, here('UAL_fleet.feather'))
```

Merging Airfleet data into our dataset

```{r, eval = FALSE}
# Load Files ----
maindf <- read_feather(here('clean_3airlines.feather'))

delta <- read_feather(here('DAL_fleet.feather'))

american <- read_feather(here('AA_fleet.feather'))

united <- read_feather(here('UAL_fleet.feather'))

# Join Files ----

allrows <- bind_rows(delta, american, united)

testjoin <- left_join(maindf, allrows, by = c('tail_num' = 'Tail'))

col_status <- status(testjoin)

# Clean column names ----
clean_join <- clean_names(testjoin)

# Separate the age/year columns into two columns
separate <- clean_join %>% 
  separate(age_built, c('age', 'year'))

status(separate)
```

Airline On-time data 
```{r, eval = FALSE}
ontime_raw <- read.csv("https://raw.githubusercontent.com/charlottetse33/36103_Logistics-Transport-Aviation/main/test%20on%20time%20report.csv")
```
Airline DOT_ID corresponding to Airline Names
```{r, eval = FALSE}
#Seperate loading data to prevent R from crashing
id_airline <- read.csv("https://raw.githubusercontent.com/charlottetse33/36103_Logistics-Transport-Aviation/main/dot_id_airline.csv")

```
Merge Airline Names with On-time Dataset 
```{r, eval = FALSE}
colnames(id_airline) <- c("DOT_ID_Reporting_Airline", "Airline")
  
ontime_1 <- left_join(ontime_raw, id_airline, by = "DOT_ID_Reporting_Airline")

```
Merge API Airport names with On-time Dataset 
```{r, eval = FALSE}
ontime_2 <- ontime_1 %>% mutate(Origin_airportCode = paste("K",Origin, sep = ""), Dest_airportCode = paste("K",Dest, sep = ""))

Originjoint <- left_join(ontime_2, API_data, by = c("Origin_airportCode" = "airportCode"))
names(Originjoint)[names(Originjoint) == "airportName"] <- "Origin_airportName"

ontime_final <- left_join(Originjoint, API_data, by = c("Dest_airportCode" = "airportCode"))
names(ontime_final)[names(ontime_final) == "airportName"] <- "Dest_airportName"

#New column names from merging data: 
#Airline, Origin_airportName, Dest_airportName

```
Adding fleet data into ontime dataset 
```{r, eval = FALSE}
maindf <- read_feather(here('clean_3airlines.feather'))

delta <- read_feather(here('DAL_fleet.feather'))

american <- read_feather(here('AA_fleet.feather'))

united <- read_feather(here('UAL_fleet.feather'))

# Join Files ----

allrows <- bind_rows(delta, american, united)

testjoin <- left_join(maindf, allrows, by = c('tail_num' = 'Tail'))

col_status <- status(testjoin)

# Clean column names ----
clean_join <- clean_names(testjoin)

# Separate the age/year columns into two columns
separate <- clean_join %>% 
  separate(age_built, c('age', 'model_year'))

status(separate)

# Save a feather file

write_feather(separate, here('clean_3airline_fleet.feather'))

```

Adding employee data into ontime dataset
```{r, eval = FALSE}
allmonths <- read_feather("C:/Users/tsetc/Downloads/clean_3airline_fleet (1).feather")
#This dataset includes airline, origin_airportname, dest_airportname 

names(allmonths) <- toupper(names(allmonths))

# read employee dataset 
employee_data <- read_csv("https://raw.githubusercontent.com/timothywallaby/36103_Logistics-Transport-Aviation/main/Employment%20dataset%202019-2021.csv")

# concentenate variables for employee dataset 
employee_data$combine <- paste(employee_data$YEAR, employee_data$MONTH, employee_data$CARRIER_NAME, sep = "_")
employee_data <- employee_data %>% select(combine, EMPFULL, EMPPART, EMPTOTAL, EMPFTE)

# concentenate variables for big dataset
allmonths$combine <- paste(allmonths$YEAR, allmonths$MONTH, allmonths$AIRLINE, sep = "_")

# Combine 2 dataset 
allmonths <- left_join(allmonths, employee_data, by = 'combine')

allmonths <- allmonths %>% select(-combine)

write_feather(allmonths, "C:/Users/tsetc/Downloads/clean_3airline_fleet_employ.feather")
```
Clean data 



One or more appendices are the place to out details and ancillary materials.
These might include such items as
• Technical descriptions of (unusual) statistical procedures
• Detailed tables or computer output
• Figures that were not central to the arguments presented in the body of the report
• Computer code used to obtain results.
In all cases, and especially in the case of computer code, it is a good idea to add some text sentences
as comments or annotations, to make it easier for the uninitiated reader to follow what you are doing.
It is often difficult to find the right balance between what to put in the appendix and what to put in
the body of the paper. Generally you should put just enough in the body to make the point, and refer
the reader to specific sections or page numbers in the appendix for additional graphs, tables and other
details.
